---
title: "Pauses in a fast paced life: Intermittent hovering in hummingbirds"
author: "Kristiina Hurme, Ana Melisa Fernandes, Diego Andrés Benitez Duarte, Marialejandra Castro-Farías, Gary Stiles, Ashley Smiley, Christopher J. Clark, Alejandro Rico-Guevara"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

## 1. Introduction and Dependencies

This document contains the reproducible workflow for the analysis of evolutionary correlations between flight behavior and morphology in hummingbirds.

```{r libraries}
# INSTALLATION AND LOADING
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!require("ggtree", quietly = TRUE)) BiocManager::install("ggtree")
if (!require("pacman")) install.packages("pacman")
if (!require("here")) install.packages("here") # Vital para rutas relativas

pacman::p_load(
    phytools, ggtree, caper, tidyverse, RColorBrewer, readxl, janitor, nlme, 
    ape, phylolm, cowplot, caret, kableExtra, ggplotify, here
)
```

## 2. Auxiliary Functions

Custom functions defined for coefficient extraction, forest plot generation, and metric calculation.

```{r functions}
source(here::here("R", "functions.R"))
```

## 3. Data Loading and Curation

Data and phylogeny are loaded using relative paths for reproducibility.

```{r load_data}
data_path <- here("Data", "Trochilidae_MorphoBehavior_Data.xlsx")
tree_path <- here("Data", "arbol.nwk")

data_raw <- read_excel(data_path, sheet = "Pausas_Tipo")
data_raw <- data_raw[data_raw$Species != "Sternoclyta cyanopectus", ]
tree_raw <- read.tree(tree_path) 

# Data Cleaning
data_raw[, c(FALSE, sapply(data_raw[,-1], is.character))] <- 
    lapply(data_raw[, c(FALSE, sapply(data_raw[,-1], is.character))], factor)
data_clean <- data_raw %>% clean_names()

data_clean$species <- gsub(" ", "_", data_clean$species)
tree_raw$tip.label <- gsub(" ", "_", tree_raw$tip.label)
```

### 3.1 Phylogenetic Tree Modification

Manual addition of missing species based on literature/taxonomy.

**Justification for tree manipulation**
The species absent from the original phylogeny was grafted onto the clade to which it belongs (e.g., node 176). A short branch length was applied to treat the divergence as a recent speciation event or a conserved change, thus ensuring the inclusion of the species' morphological data in the analysis without overestimating its evolutionary uniqueness and without significantly affecting the phylogeny's topology. Values were sought that would not significantly alter the outcome of a PGLS or a phylogenetic signal (Lambda), aiming for small values compared to the total height of the tree.

```{r tree_mod}
tree_mod <- bind.tip(tree_raw, "Aglaeactis_aliciae", edge.length = 0.2, where = 176)
tree_mod <- bind.tip(tree_mod, "Lamprolaima_rhami", edge.length = 3.2, where = 107)
tree_mod <- bind.tip(tree_mod, "Sternoclyta_cyanopectus", edge.length = 6.0, where = 108)
tree_mod <- bind.tip(tree_mod, "Oxypogon_lindenii", edge.length = 1.6, where = 230)
tree_mod <- bind.tip(tree_mod, "Urochroa_leucura", edge.length = 0.8, where = 154)

# Taxonomy Harmonization
name_replacements <- c(
    "Chlorostilbon_notatus" = "Chlorestes_notata", 
    "Damophila_julie" = "Chlorestes_julie",
    "Goethalsia_bella" = "Goldmania_bella", 
    "Lophornis_pavoninus" = "Lophornis_adorabilis",
    "Cyanophaia_bicolor" = "Riccordia_bicolor", 
    "Amazilia_beryllina" = "Saucerottia_beryllina",
    "Amazilia_castaneiventris" = "Saucerottia_castaneiventris", 
    "Amazilia_cyanura" = "Saucerottia_cyanura",
    "Heliodoxa_cyanopectus" = "Sternoclyta_cyanopectus", 
    "Campylopterus_villaviscencio" = "Campylopterus_villaviscensio"
)

tree_mod$tip.label <- ifelse(tree_mod$tip.label %in% names(name_replacements), 
                             name_replacements[tree_mod$tip.label], tree_mod$tip.label)

if (any(tree_mod$edge.length <= 0)) tree_mod$edge.length[tree_mod$edge.length <= 0] <- 1e-6
if (!is.ultrametric(tree_mod)) tree_mod <- chronopl(tree_mod, lambda = 1)
tree_mod$tip.label <- gsub(" ", "_", tree_mod$tip.label)

# Final Match
common_spp <- intersect(data_clean$species, tree_mod$tip.label)
tree_bin <- keep.tip(tree_mod, common_spp)
data_bin <- data_clean %>% filter(species %in% common_spp) %>% arrange(match(species, tree_bin$tip.label))
data_bin <- as.data.frame(data_bin)
rownames(data_bin) <- data_bin$species

data_bin <- data_bin %>%
    filter(hovering_pauses %in% c("YES", "NO")) %>%
    filter(color %in% c("YES", "NO")) %>%
    filter(!is.na(wing_length)) %>% 
    mutate(winglength_vs_mass = winglength_rel_mass2)

final_spp <- intersect(data_bin$species, tree_bin$tip.label)
tree_bin <- keep.tip(tree_bin, final_spp)
data_bin <- data_bin[match(tree_bin$tip.label, data_bin$species), ]

if (any(duplicated(tree_bin$tip.label))) {
    dup_idx <- which(duplicated(tree_bin$tip.label))
    tree_bin <- drop.tip(tree_bin, dup_idx)
    data_bin <- data_bin[match(tree_bin$tip.label, data_bin$species), ]
}

data_bin$hovering_pauses_01 <- ifelse(data_bin$hovering_pauses == "YES", 1, 0)
levels_bin <- c("NO", "YES")
cols_bin <- RColorBrewer::brewer.pal(3, "Dark2")[1:2]
names(cols_bin) <- levels_bin
```

## 4. Phylogenetic Signal & Modeling

### 4.1 Phylogenetic Signal (D Statistic)

```{r phylo_signal}
signals_bin <- list()
for (cat in levels_bin) {
    var_bin <- as.numeric(data_bin$hovering_pauses == cat)
    names(var_bin) <- data_bin$species
    data_temp <- data.frame(species = data_bin$species, state_bin = var_bin, stringsAsFactors = FALSE)
    comp_data <- comparative.data(tree_bin, data_temp, names.col = "species")
    d_res <- phylo.d(comp_data, binvar = state_bin, permut = 100)
    signals_bin[[cat]] <- d_res
}
print(signals_bin)
```

### 4.2 PhyloGLM & Heatmap

```{r pgls_model, fig.width=8, fig.height=8}
model_hover_yes <- phyloglm(hovering_pauses_01 ~ wing_length, 
                            data = data_bin, phy = tree_bin, method = "logistic_MPLE")

# Probability Heatmap
prob_yes <- model_hover_yes$fitted.values
prob_matrix <- cbind(NO = 1 - prob_yes, YES = prob_yes)
rownames(prob_matrix) <- data_bin$species

if (all(!is.na(prob_matrix))) {
    phylo.heatmap(tree_bin, prob_matrix, standardize = FALSE, 
                  legend = TRUE, fsize = 0.6, colors = c("#E41A1C", "#377EB8"))
    title("Predicted Probabilities (P(NO), P(YES))")
}
```

## 5. Evolutionary Correlations (Pagel)

Testing correlated evolution between pauses and microhabitat/foraging strategy.

```{r pagel}
# Correlation vs Microhabitat
data_pagel_hab <- data_bin %>%
    mutate(habitat_bin = case_when(
        microhabitat %in% c("Canopy", "Understory") ~ "Closed",
        microhabitat %in% c("Open", "Mixed")        ~ "Open",
        TRUE ~ NA_character_ 
    )) %>%
    filter(!is.na(habitat_bin), !is.na(hovering_pauses)) %>% droplevels()

sp_hab <- intersect(data_pagel_hab$species, tree_bin$tip.label)
tree_hab <- keep.tip(tree_bin, sp_hab)
data_pagel_hab <- data_pagel_hab[match(sp_hab, data_pagel_hab$species), ]

r1 <- setNames(factor(data_pagel_hab$hovering_pauses), data_pagel_hab$species)
r2 <- setNames(factor(data_pagel_hab$habitat_bin), data_pagel_hab$species)

if (length(levels(r1)) == 2 && length(levels(r2)) == 2){
    fit_dep_hab <- fitPagel(tree_hab, r1, r2, model = "ER")
    print(fit_dep_hab)
}

# Correlation vs Foraging
data_pagel_for <- data_bin %>%
    mutate(forage_bin = case_when(
        foraging_type == "Territorial" ~ "Territorial",
        foraging_type == "Opportunist" ~ "No_Territorial",
        TRUE ~ NA_character_ 
    )) %>%
    filter(!is.na(forage_bin), !is.na(hovering_pauses)) %>% droplevels()

sp_for <- intersect(data_pagel_for$species, tree_bin$tip.label)
tree_for <- keep.tip(tree_bin, sp_for)
data_pagel_for <- data_pagel_for[match(sp_for, data_pagel_for$species), ]
r3 <- setNames(factor(data_pagel_for$forage_bin), data_pagel_for$species)

if (length(levels(r1)) == 2 && length(levels(r3)) == 2){
    fit_dep_for <- fitPagel(tree_for, r1, r3, model = "ER")
    print(fit_dep_for)
}
```

## 6. Visualization

### 6.1 Tanglegram: Pauses vs Coloration

```{r tanglegram, fig.width=12, fig.height=10}
# Colors and Clades 
tips_in_order <- tree_bin$tip.label
clades_ordered <- data_bin$clade[match(tips_in_order, data_bin$species)]
unique_clades <- unique(rev(clades_ordered))
cols_clades <- RColorBrewer::brewer.pal(max(length(unique_clades), 3), "Paired")[1:length(unique_clades)]
names(cols_clades) <- unique_clades

edge_cols <- rep("gray10", nrow(tree_bin$edge))
edge_widths <- rep(1, nrow(tree_bin$edge))
for (c in unique_clades) {
    sp_clade <- intersect(data_bin$species[data_bin$clade == c], tree_bin$tip.label)
    if (length(sp_clade) > 0) {
        idx <- which.edge(tree_bin, sp_clade)
        edge_cols[idx] <- cols_clades[c]
        edge_widths[idx] <- 4
    }
}

wing_levels <- unique(trimws(tolower(as.character(data_bin$wing_color))))
cols_wing_map <- setNames(wing_levels, wing_levels)
if ("indigo" %in% names(cols_wing_map)) cols_wing_map["indigo"] <- "#4B0082"
if ("sapphire" %in% names(cols_wing_map)) cols_wing_map["sapphire"] <- "#2138AB"
if ("white" %in% names(cols_wing_map)) cols_wing_map["white"] <- "lightgoldenrod2"
if ("gray" %in% names(cols_wing_map)) cols_wing_map["gray"] <- "gray35"
if ("orange" %in% names(cols_wing_map)) cols_wing_map["orange"] <- "darkorange"

ordered_idx <- match(tree_bin$tip.label, data_bin$species)
ordered_colors <- as.character(data_bin$wing_color[ordered_idx])
cols_wing_sp <- cols_wing_map[trimws(tolower(ordered_colors))]

max_depth <- max(node.depth.edgelength(tree_bin))
xlim_l <- c(-max_depth * 0.35, max_depth * 2.6) 
xlim_r <- c(-max_depth * 0.35, max_depth * 2.6)

clean_labels <- gsub("_", " ", tree_bin$tip.label)
clean_labels <- gsub("Coeligena bonapartei", "Coeligena bonapartei/consita", clean_labels)

layout(matrix(c(1,2), 1, 2))
par(oma = c(0, 0, 0, 0), mar = c(0, 0, 2, 0)) 

# Left: Hovering Pauses
states_hover <- setNames(as.character(data_bin$hovering_pauses), data_bin$species)
ace_hover <- ace(states_hover, tree_bin, type = "discrete", model = "ER")
cols_hover_fix <- c("NO" = "#1B9E77", "YES" = "#D95F02") 

plot(tree_bin, cex = 0.8, no.margin = TRUE, label.offset = 0.02, 
     direction = "rightwards", show.tip.label = FALSE, 
     x.lim = xlim_l, edge.color = edge_cols, edge.width = edge_widths)
nodelabels(pie = ace_hover$lik.anc, piecol = cols_hover_fix, cex = 0.45)
tiplabels(pch = 21, bg = cols_hover_fix[states_hover], cex = 2)
legend("bottomleft", legend = names(cols_hover_fix), fill = cols_hover_fix,
       title = "Hovering Pauses", cex = 1, bty = "n")

# Center: Labels
pos_center <- max_depth * (2.6 - 1) * 0.5 + (max_depth * 0.1)
tiplabels(text = clean_labels, frame = "none", col = cols_wing_sp, 
          font = 2, cex = 0.8, adj = 0.5, offset = pos_center)

# Right: Underwing Coloration
if (!exists("ace_color")) {
    states_col <- setNames(as.character(data_bin$color), data_bin$species)
    ace_color <- ace(states_col, tree_bin, type = "discrete", model = "ER")
}
plot(tree_bin, cex = 0.8, no.margin = TRUE, label.offset = 0.02, 
     direction = "leftwards", show.tip.label = FALSE, 
     x.lim = xlim_r, edge.color = edge_cols, edge.width = edge_widths)
cols_right <- c("NO" = "gray80", "YES" = "#4B0082")

nodelabels(pie = ace_color$lik.anc, piecol = cols_right, cex = 0.45)
tiplabels(pch = 21, bg = cols_right[data_bin$color], cex = 2)
legend("bottomright", legend = c("Gray", "Colorful"), 
       fill = c("gray80", "#4B0082"), title = "Underwing Coloration", cex = 1, bty = "n")
legend("topright", legend = names(cols_clades), col = cols_clades, 
       lwd = 5, title = "Clades", cex = 0.8, bty = "n", ncol = 1)
par(mfrow = c(1,1))
```

### 6.2 Composite Plot: Violins + Forest Plots

```{r composite_plot, fig.width=10, fig.height=8}
cols_hover_plt <- RColorBrewer::brewer.pal(3, "Dark2")[1:2]
names(cols_hover_plt) <- c("NO", "YES")
cols_color_plt <- c("NO" = "gray40", "YES" = "#4B0082")

plot_data <- data_bin %>%
    filter(hovering_pauses %in% c("YES", "NO"),
           color %in% c("YES", "NO"),
           !is.na(wing_length)) %>% droplevels()

# Violins
p_violin_hover <- ggplot(plot_data, aes(x = hovering_pauses, y = wing_length, fill = hovering_pauses)) +
    geom_violin(trim = FALSE, alpha = 0.6, color = NA) +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, alpha = 0.8) +
    geom_jitter(width = 0.1, size = 1.5, alpha = 0.4) +
    scale_fill_manual(values = cols_hover_plt) +
    labs(title = "Wing Length vs Hovering Pauses", y = "Wing Length (mm)", x = "Hovering Pauses") +
    theme_minimal() + theme(legend.position = "none", plot.title = element_text(face = "bold", hjust = 0.5, size = 12))

p_violin_color <- ggplot(plot_data, aes(x = color, y = wing_length, fill = color)) +
    geom_violin(trim = FALSE, alpha = 0.6, color = NA) +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, alpha = 0.8) +
    geom_jitter(width = 0.1, size = 1.5, alpha = 0.4) +
    scale_fill_manual(values = cols_color_plt) +
    scale_x_discrete(labels = c("NO" = "Gray", "YES" = "Colorful")) +
    labs(title = "Wing Length vs Underwing Coloration", y = "Wing Length (mm)", x = "Underwing Coloration") +
    theme_minimal() + theme(legend.position = "none", plot.title = element_text(face = "bold", hjust = 0.5, size = 12))

# Forest Plots
df_conf_hover <- get_forest_data("hovering_pauses", "wing_length", data_bin, tree_bin, cols_hover_plt)
df_conf_color <- get_forest_data("color", "wing_length", data_bin, tree_bin, cols_color_plt)

p_conf_hover <- create_forest_plot(df_conf_hover, cols_hover_plt, "Hovering Pauses")
p_conf_color <- create_forest_plot(df_conf_color, cols_color_plt, "Underwing Coloration")

plot_grid(p_conf_hover, p_violin_hover, p_conf_color, p_violin_color, ncol = 2, align = 'h', labels = c("A", "B", "C", "D"))
```

### 6.3 Clade-Level Morphology

```{r clade_plot, fig.width=12, fig.height=12}
df_phylo_morpho <- data_clean %>%
    filter(species != "Sternoclyta_cyanopectus", !is.na(wing_length)) %>% droplevels()

tree_phylo_morpho <- keep.tip(tree_mod, intersect(tree_mod$tip.label, df_phylo_morpho$species))
df_phylo_morpho <- df_phylo_morpho[match(tree_phylo_morpho$tip.label, df_phylo_morpho$species), ]

df_phylo_morpho$clean_label <- gsub("_", " ", df_phylo_morpho$species)
df_phylo_morpho$clean_label <- gsub("Coeligena bonapartei", "Coeligena bonapartei/consita", df_phylo_morpho$clean_label)

# Map colors
df_phylo_morpho$tip_color_hex <- cols_wing_map[trimws(tolower(as.character(df_phylo_morpho$wing_color)))]
na_idx <- which(is.na(df_phylo_morpho$tip_color_hex))
if(length(na_idx) > 0) df_phylo_morpho$tip_color_hex[na_idx] <- "black" 
white_idx <- which(df_phylo_morpho$tip_color_hex %in% c("white", "#FFFFFF", "#ffffff"))
if(length(white_idx) > 0) df_phylo_morpho$tip_color_hex[white_idx] <- "gray40"

# Legend
legend_data <- data.frame(Color_Name = names(cols_wing_map), Color_Hex = as.character(cols_wing_map))
legend_data$Color_Name[legend_data$Color_Hex == "beige"] <- "white"
p_legend <- ggplot(legend_data, aes(x=1, y=1, color=Color_Name)) +
    geom_point(size=4) +
    scale_color_manual(name = "Wing Coloration", values = cols_wing_map) +
    theme_void() +
    theme(legend.position = "bottom", legend.title = element_text(face="bold"), legend.text = element_text(size=10))
wing_legend <- get_legend(p_legend)

# Tree
clade_list <- split(df_phylo_morpho$species, df_phylo_morpho$clade)
tree_grouped_full <- groupOTU(tree_phylo_morpho, clade_list)

p_tree_final <- ggtree(tree_grouped_full, aes(color = group), layout = "rectangular", size = 1.1) %<+% df_phylo_morpho +
    scale_color_manual(values = c("0" = "black", cols_clades)) +
    geom_tiplab(aes(label = clean_label, color = I(tip_color_hex)), size = 2.8, fontface = "bold", align = TRUE, linesize = 0.3, offset = 0.02) +
    theme_tree() + 
    theme(legend.position = "none", plot.margin = margin(10, 0, 10, 10)) + 
    xlim(0, max(node.depth.edgelength(tree_phylo_morpho)) * 1.4)

# Violin
tips_order_final <- get_taxa_name(p_tree_final)
ordered_df <- df_phylo_morpho[match(tips_order_final, df_phylo_morpho$species), ]
clade_appearance <- unique(ordered_df$clade)
df_phylo_morpho$clade_ordered <- factor(df_phylo_morpho$clade, levels = rev(clade_appearance))

p_violin_final <- ggplot(df_phylo_morpho, aes(x = clade_ordered, y = wing_length, fill = clade_ordered)) +
    geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA) +
    geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA, alpha = 0.8) +
    geom_jitter(width = 0.1, size = 1, alpha = 0.4, color = "black") +
    scale_fill_manual(values = cols_clades) +
    coord_flip() +
    labs(title = "Wing Length Distribution", x = "", y = "Wing Length (mm)") +
    theme_minimal() +
    theme(legend.position = "none", axis.text.y = element_text(size = 10, face = "bold", color = "black"),
          panel.grid.major.y = element_blank())

main_plot <- plot_grid(p_tree_final, p_violin_final, ncol = 2, align = "h", axis = "bt", rel_widths = c(1.8, 1), labels = c("A", "B"))
plot_grid(main_plot, wing_legend, ncol = 1, rel_heights = c(1, 0.1))
```

## 7. Comparative Summary Table

Summary of evolutionary statistics and PGLS models.

```{r summary_table}
res_hover <- calculate_full_metrics("hovering_pauses", data_bin, tree_bin)
res_trans <- calculate_full_metrics("transition_pauses", data_bin, tree_bin)
res_color <- calculate_full_metrics("color", data_bin, tree_bin)

summary_table <- data.frame(
    Metric = names(res_hover),
    Hovering = res_hover,
    Transitions = res_trans,
    Coloration = res_color,
    row.names = NULL, stringsAsFactors = FALSE
)

# Highlighting significant results
idx_pvals <- c(7, 11, 15)
red_rows <- c()
for (i in idx_pvals) {
    if (as.numeric(summary_table[i, "Hovering"]) < 0.05 & 
        as.numeric(summary_table[i, "Transitions"]) < 0.05 & 
        as.numeric(summary_table[i, "Coloration"]) < 0.05) {
        red_rows <- c(red_rows, i)
    }
}

kbl_table <- summary_table %>%
    kbl(caption = "Comparative Summary of Evolutionary Statistics and PGLS Models", align = "c") %>%
    kable_classic(full_width = F, html_font = "Cambria") %>%
    row_spec(0, bold = T, color = "white", background = "#2c3e50") %>%
    column_spec(1, bold = T, border_right = T, width = "4cm", color = "#555") %>%
    pack_rows("1. Evolutionary Signal (Categorical Trait)", 1, 4, label_row_css = "background-color:#ecf0f1;color:#333;font-weight:bold;") %>%
    pack_rows("2. Model: Wing Length ~ Trait", 5, 8, label_row_css = "background-color:#d1f2eb;color:#0e6655;font-weight:bold;") %>%
    pack_rows("3. Model: Body Mass ~ Trait", 9, 12, label_row_css = "background-color:#d6eaf8;color:#154360;font-weight:bold;") %>%
    pack_rows("4. Model: Relation Wing Length vs Body Mass ~ Trait", 13, 16, label_row_css = "background-color:#fce8d6;color:#a04000;font-weight:bold;") %>%
    row_spec(idx_pvals, bold = T)

if (length(red_rows) > 0) kbl_table <- kbl_table %>% row_spec(red_rows, color = "red", bold = T)
kbl_table
```